Emittance preservation of an electron beam in a loaded quasi-linear plasma wakefield

### 2017-10-08 Tex/Paper.tex
Implemented changes from Kyrre

Beam driven plasma wakefield accelerators \cite{chen:1985} have the potential to offer compact linear accelerators with
high energy gradients, and have been of interest for several decades. With a relativistic charged particle drive beam
travelling through a plasma, a strong wakefield is excited that can be loaded by a trailing witness beam. When the
witness beam optimally loads the wakefield, an increase in absolute energy spread can be kept small. The concept has
been demonstrated experimentally in the past using electron drive beams accelerating electron witness beams
\cite{rosenzweig:1988, blumenfeld:2007, kallos:2008, litos:2014}.

A major challenge with plasma wakefield accelerators is, however, to accelerate a beam while keeping both energy spread
and emittance growth small. In the well described linear regime, valid when the beam density $n_{b}$ is much smaller
than the plasma density $n_{0}$, a non-linear transverse focusing force causes emittance growth of the witness beam.
The beam also sees a transversely and longitudinally varying accelerating field causing a spread in energy after the
beam has been accelerated \cite{katsouleas:1987}. In the non-linear regime, where $n_{b} > n_{0}$, a bubble is formed by
the transverse oscillations of the plasma electrons, gathering in a sheath around an evacuated area filled with only
ions. The ions, assumed stationary, form a uniform density ion channel creating a focusing force that varies linearly
with radius. This focusing force preserves emittance \cite{rosenzweig:1991}.

In this paper we present simulation results showing how emittance preservation of a high charge density witness beam can
be ensured when accelerated by a proton drive beam producing quasi-linear wakefields \cite{rosenzweig:2010}. By
quasi-linear wakefields we here mean wakefields with only partial blow out of the plasma electrons in the bubble.
The key idea is to have enough charge in the witness beam to at the same time load the wakefield to produce low relative
energy spread, and completely blow out the electrons left in the bubble after the beam to reach conditions that
preserves emittance. The results have importance for the preparation of AWAKE Run 2 \cite{adli:2016}, and possibly other
applications in the quasi-linear regime.

@@

% 	EA comment: is this part relevant?  Feels out of place.  We repeat this physics below when discussing proton beam SMI.  
%A train of electron drive bunches with a separation $\lambda_{pe}$ and a length $l_{b} \ll \lambda_{pe}$, where the
%plasma wavelength $\lambda_{pe} = 2\pi c/\omega_{pe}$ and the plasma frequency $\omega_{pe} = (n_{0} e^{2} / m_{e}
%\epsilon_{0})^{1/2}$, produces a field $E_{z}$ that increases for each drive bunch \cite{chen:1985}. A tailing witness
%beam loading the peak accelerating phase of this field will gain energy from the wakefield from all of the drive bunches
%when they are properly phased with respect to the wakefields. Acceleration of an electron witness beam driven by two
%electron drive beams has been demonstrated at Brookhaven National Laboratory \cite{muggli:2011}.

@@

\begin{figure}[hbt]
    \includegraphics[width=0.99\linewidth,trim={1mm 2mm 1mm 2mm},clip]{figures/figAWAKE}
    \caption{\label{Fig:AWAKER2} A simplified illustration of the experimental setup for AWAKE Run 2. The SPS proton
        beam undergoes self-modulation in the first plasma section. The electron witness beam is injected into the
        accelerating structure, and undergoes acceleration in the second plasma section
        \cite{berglyd_olsen:2015, adli:2016}}.
\end{figure}

The energy carried by electron drive beams used in previous plasma wakefield experiments have been on the order of
$100\unit{J}$ and the propagation length typically no more than $1\unit{m}$ \cite{blumenfeld:2007, caldwell:2009}.
For high-energy physics application a higher total beam energy is often desired. For instance, the energy of a
high-charge electron beam accelerated to $1\unit{TeV}$ with $1\nexp{10}$ electrons, similar to the beam that could be
produced by the International Linear Collider, is $1.6\unit{kJ}$. Using electron beams as drivers, a large number of
plasma stages is required to reach an energy of a $\unit{kJ}$ for the accelerated beam. However, staging plasma
accelerators without reducing the effective gradient and spoiling the beam quality is challenging
\cite{steinke:2016, lindstrom:2016}.

Proton beams available at CERN carry significantly more energy than available electron beams, $19\unit{kJ}$ for the SPS
beam \cite{gschwendtner:2016}, allowing for much longer plasma wakefield accelerator stages. The SPS beam is orders of
magnitude longer than the plasma wavelengths needed for such applications, and it does not drive a strong wake
\cite{gschwendtner:2016}. By letting the proton beam undergo self-modulation before injecting the witness beam into the
accelerating structure, stronger wakefields are excited. The self-modulation is produced by the transverse fields
generated by the beam acting upon itself, causing regions of the beam to rapidly defocus \cite{kumar:2010}. The
modulation frequency is close to that of the plasma, and produces a train of short proton bunches along the beam axis
with a surrounding halo of defocused particles. This train of bunches resonantly drives wakefields to large amplitudes.

AWAKE at CERN is a proof of concept proton beam driven plasma wakefield accelerator experiment
\cite{awake_collaboration:2014}, currently in its first phase of operation. The experiment uses a $400\unit{GeV}$ proton
beam delivered by the SPS as its driver, and a single $10\unit{m}$ plasma stage with a nominal plasma density of
$7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. This plasma density corresponds to $\lambda_{pe} = 1.26\unit{mm}$
-- matched to the transverse size of the SPS proton beam such that $k_{pe}\sigma_{x,y,pb} = 1$, where
$k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave number, $\lambda_{pe} = 2\pi c/\omega_{pe}$, and
$\omega_{pe} = \left(n_0e^2/m_e\epsilon_0\right)^{1/2}$ is the plasma electron angular frequency.

The aim of the first phase of the experiment is to demonstrate self-modulation of the proton beam. The aim of the second
phase, in 2018, is to sample the wakefield with a long electron beam $\simeq\lambda_{pe}$. The study presented here has
relevance for Run 2 \cite{adli:2016}, starting in 2021 after the LHC long shutdown 2, and aims to demonstrate
acceleration of a short electron beam, $\ll\lambda_{pe}$, to high energy and with a minimal increase in emittance and
absolute energy spread.

The plans for AWAKE Run 2 proposes to use two plasma sections, illustrated in Fig. \ref{Fig:AWAKER2}. The first
section of about $4\unit{m}$ is the self-modulation stage where the proton beam undergoes self-modulation without the
electron beam present. The electron witness beam is then injected into the modulated proton beam before section two
where it undergoes acceleration. The self-modulated proton beam does not produce a fully non-linear wakefield, and
therefore not all plasma electrons are evacuated from the plasma bubble. The result is that the focusing force does not
increase linearly with radius and the accelerated beam emittance is not preserved.

@@

The focus in this study is on the beam loading of the wakefields driven by the proton beam. Studies of self-modulated
proton beams show that the proton beam evolves as it propagates through a uniform plasma \cite{lotov:2011}, but that
small variations in the plasma profile the modulation, and thus the wakefields, may be stabilised over long distances
\cite{lotov:2011, lotov:2015, caldwell:2011}. To study the witness beam evolution in a stable wake, independent of the
dynamics of the self-modulation, we use a single, non-evolving proton bunch as driver. The proton beam parameters
are chosen so that key features of the wake -- the plasma electron density in the wake and the longitudinal electric
field -- are the same as in the wake of a self-modulated proton beam with AWAKE baseline parameters
\cite{gschwendtner:2016}. Both the proton beam and the witness beam have Gaussian longitudinal charge distribution and
bi-Gaussian transverse charge distributions.

We have previously studied the beam loading in a proton beam wake using the full particle-in-cell (PIC) code OSIRIS
\cite{fonseca:2002} with 2D cylindrical-symmetric simulations. The studies \cite{berglyd_olsen:2015, berglyd_olsen:2016}
primarily looked at beam loading, energy gain and energy spread, as well as different approaches to creating a stable
drive beam structure based on previous self-modulation studies. In order to study the witness beam emittance evolution
we use the recently released open-source version of QuickPIC \cite{huang:2006, an:2013}. QuickPIC is a fully
relativistic 3D quasi-static PIC code. It does not suffer from the numerical Cherenkov effect that full PIC codes do
\cite{godfrey:1974,lehe:2013}, making it a well suited tool to study emittance preservation. All simulation results in
this paper were obtained using QuickPIC open-source \cite{quickpic:web}.

@@

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/plasmaDenTWake}
    \caption{\label{Fig:PlasmaDenTWake} QuickPIC simulations of the initial time step for the single proton drive beam
        and witness beam setup. Plasma electron density is shown in grey with the drive beam (blue) and the witness beam
        (red) superimposed. The line plot indicates the transverse wakefield gradient $dW_{x}/dx$ where
        $W_{x} = E_{x} - v_{b} B_{y}$, evaluated along the beam axis. Beams move to the left.}
\end{figure}

@@

The modulation process used in AWAKE does not reach the fully non-linear regime and thus does not produce a bubble void
of plasma electrons. When the SPS beam, containing $3\nexp{11}$ protons \cite{gschwendtner:2016}, enters the second
plasma section (Fig. \ref{Fig:AWAKER2}), the peak electric field is expected to be about $500\funit{MV}{m}$. The
plasma electron density is only depleted to around $65\%$ of nominal plasma density at the point where we inject the
electron beam \cite{awake_collaboration:2016}. The plasma electron density depletion and the peak field is replicated
using a single bunch with $1.46\nexp{10}$ protons ($2.34\unit{nC}$), a length $\sigma_{z} = 40\unit{\mu m}$
($7\unit{kA}$), and a transverse size $\sigma_{x,y,pb} = 200\unit{\mu m}$. The peak density at $0.83\cdot n_{0}$ results
in a quasi-linear wake. To avoid transverse evolution of the proton driver, emulating the stable propagation of the
self-modulated proton beam \cite{lotov:2011, lotov:2015, caldwell:2011}, we freeze the transverse evolution of the
equivalent proton bunch by increasing the particles mass by six orders of magnitude.

@@

In order to prevent large amplitude oscillations of the witness beam particles, which may cause additional energy spread
as well as emittance growth, we consider a witness beam matched to the plasma density. The matched beam transverse size
is:
\begin{equation}
    \sigma_{x,y,eb}=\left(\frac{2c^{2}\epsilon_{\mathrm{N}}^{2}m_{e}\varepsilon_{0}}{n_{pe}e^{2}\gamma}\right)^{1/4}
    \label{EQ:Matched}
\end{equation}

@@

Beam loading of a short witness beam is sensitive to its position relative to the electric field \cite{tzoufras:2009}
as well as to its de-phasing with respect to the wakefields at low energy. To eliminate de-phasing of the witness beam,
the initial beam energy is set such that $\gamma_{eb} = \gamma_{pb} = 426.3$, giving an energy of $217\unit{MeV}$. A
lower initial energy is likely to be sufficient for AWAKE Run 2 injection.
% EA: we have studied this to some extent, but unfortunately we do not have a good refence for the last statement,
%     no details in \cite{berglyd_olsen:2015}.

Equation~(\ref{EQ:Matched}) yields a transverse size $\sigma_{x,y,eb}$ of $5.25\unit{\mu m}$, which is narrow compared
to the drive beam $\sigma_{x,y,pb} = 200\unit{\mu m}$. The bunch length was set to $\sigma_{z} = 60\unit{\mu m}$, based
on earlier beam loading studies \cite{berglyd_olsen:2016}. The charge is adjusted to $100\unit{pC}$ for optimal beam
loading, as discussed in the next section. We refer to the defined drive beam and witness beam parameter set as the
\emph{base case}. Figure \ref{Fig:PlasmaDenTWake} shows the two beams from a QuickPIC simulation of the initial time
step, for the base case parameters; the proton beam in blue, the trailing electron beam in red, and the plasma electron
density in grey.

@@

The relatively small size of the witness beam puts constraints on the transverse grid cell size and number in the
simulations. We need a small size to resolve the narrow electron beam, and a large number of grid cells to resolve the
much wider proton beam and its wakefields. We use a transverse grid cell size of $1.17\unit{\mu m}$, and of
$2.34\unit{\mu m}$ for the longitudinal grid cells for the simulations presented in section \ref{S:BL}. The witness beam
was simulated with $16.8\nexp{6}$ non-weighted particles, the drive beam with $2.1\nexp{6}$ non-weighted particles, and
the plasma electrons with $1024 \times 1024$ weighted particles per transverse slice.

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamLoading}
    \caption{\label{Fig:BeamLoading} \textbf{Top plot:} Unloaded longitudinal electric field with no witness beam
        (dashed blue line) and loaded field (whole blue line) along the beam axis. The beam density along the axis for
        both beams are shown in red.
        \textbf{Bottom plot:} Plasma densities along the beam axis for a drive beam with no witness beam (dashed green
        line), witness beam with no drive beam (dash-dotted green line), and both beams present (whole green line).
        $\xi = z - tc$ is the position in the simulation box, moving towards the left.}
\end{figure}

@@

\begin{figure*}[hbt]
    \includegraphics[width=0.495\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamEmittance}
    \includegraphics[width=0.495\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamEmittanceOffset}
    \caption{\label{Fig:BeamEmitt} Beam density in blue along the beam axis for an on-axis beam with respect to the
        drive beam axis (left), and an offset beam (right) with an offset of one $\sigma_{x,eb} = 5.24\unit{\mu m}$ in
        the x-plane -- at four different positions $z$ in the plasma stage. The red lines show a moving window
        calculation of transverse normalised emittance. The moving window calculation uses longitudinal slices of
        $l = 4\times\Delta\xi = 9.38\unit{\mu m}$ with a step of $\Delta\xi$. Only slices with more than $100$ macro
        particles have been included. The plasma density profile is included in green, and scaled up by a factor of
        $100$ to be visible. These simulations were run with an LHC energy drive beam of $7\unit{TeV}$}
\end{figure*}

@@

Figure \ref{Fig:BeamLoading} shows the results of QuickPIC simulations of the initial time step for the base case
parameters. The $E_{z}$-field generated by the proton drive beam is seen as the blue line, shown with and without the
electron beam present. With a proton beam density $n_{pb} \simeq n_{0}$, the wakefields are in the quasi-linear regime
\cite{rosenzweig:2010}. The dashed green line in the lower part of Fig. \ref{Fig:BeamLoading} shows that the on-axis
plasma density has a depletion to $67\%$, close to what we see in full scale reference simulations for AWAKE Run 2
\cite{awake_collaboration:2016}.

@@

The witness beam generates its own wakefield which loads the $E_{z}$-field generated by the drive beam. With an ideally
shaped electron beam charge profile it is possible to optimally load the field in such a way that the accelerating field
is constant along the beam \cite{katsouleas:1987, tzoufras:2009}. Gaussian beams, as assumed in these studies, cannot
completely flatten the electric field in the tails of the charge distribution, and our base case beam therefore has a
tail in energy both at the front and the back of the beam, as illustrated in Fig. \ref{Fig:BeamPS}. The bulk of the
beam, however, sees a relatively flat field.
% EA: I rewrote this to the sentence after the comments
%we can acceleratoreHowever, with such a narrow beam when matched beam to the plasma density (equation
%\ref{EQ:MatchedB}), we are limited on how much total charge we can accelerate without reaching charge densities that
%will overload the field.
%  EA: shall we say the below?  The sentence asks for more detailed studies of the dependence of the loading of the transverse fields.
%The amounts of charge required to load the field depends both on the witness bunch length as well as the transverse size.

The initial electron beam density is $n_{eb} \approx 35\cdot n_{0}$. This means that the witness beam's own wakefield
is in the fully non-linear regime, where the space charge force is sufficient to blow out all plasma electrons,
resulting in the formation of an ion column (see Fig. \ref{Fig:BeamLoading}, bottom). This ion column, as is well
known \cite{rosenzweig:1991}, provides a linear focusing force on the part the electron beam within the column, and
therefore prevents emittance growth for this part of the beam. This effect is shown for our base case in Fig.
\ref{Fig:PlasmaDenTWake}. The focusing field has a gradient of $20\unit{kT/m}$ near the beam axis, corresponding to the
matched field gradient.

@@

Figure \ref{Fig:BeamEmitt} shows the slice emittance along the beam for the base case, sampled after propagating through
$0$, $4$, $40$ and $100\unit{m}$ of plasma. We define emittance of a slice as preserved if the growth is less than
$5\%$, and $\widetilde{Q}$ as the sum charge of the slices for which the emittance is preserved. Simulation results show
that $\widetilde{Q}/Q = 73\%$ of the electron beam longitudinal slices retain their initial emittance after the
propagation in the plasma. Emittance growth mainly occurs in the first few metres, and no significant emittance growth
is observed after this for propagation lengths up to $100\unit{m}$. The head of the beam does not benefit from the full
ion column focusing, but since the proton beam creates a quasi-linear wake, the emittance of the head of the beam still
stabilises after some time. For the $100\unit{m}$ simulation, the drive beam energy was increased to $7\unit{TeV}$ (LHC
energy) to prevent de-phasing, as de-phasing starts to become a significant effect for the SPS beam of $400\unit{GeV}$
after about $50\unit{m}$.

>>>>

Beam driven plasma wakefield accelerators~\cite{chen:1985} have the potential to offer compact linear accelerators with
high energy gradients, and have been of interest for several decades. With a relativistic charged particle drive beam
travelling through a plasma, a strong wakefield is excited that can be loaded by a trailing witness beam. When the
witness beam optimally loads the wakefield, an increase in absolute energy spread can be kept small. The concept has
been demonstrated experimentally in the past using electron drive beams accelerating electron witness
beams~\cite{rosenzweig:1988, blumenfeld:2007, kallos:2008, litos:2014}.

A major challenge with plasma wakefield accelerators is, however, to accelerate a beam while keeping both energy spread
and emittance growth small. In the well described linear regime, valid when the beam density $n_{b}$ is much smaller
than the plasma density $n_{0}$, a non-linear transverse focusing force causes emittance growth of the witness beam.
The beam also sees a transversely and longitudinally varying accelerating field causing a spread in energy after the
beam has been accelerated~\cite{katsouleas:1987}. In the non-linear regime, where $n_{b} > n_{0}$, a bubble is formed by
the transverse oscillations of the plasma electrons, gathering in a sheath around an evacuated area filled with only
ions. The ions, assumed stationary, form a uniform density ion channel creating a focusing force that varies linearly
with radius. This focusing force preserves emittance~\cite{rosenzweig:1991}.

In this paper we present simulation results showing how emittance preservation of a high charge density witness beam can
be ensured when accelerated by a proton drive beam producing quasi-linear wakefields~\cite{rosenzweig:2010}. By
quasi-linear wakefields we here mean wakefields with only partial blow out of the plasma electrons in the bubble.
The key idea is to have enough charge in the witness beam to at the same time load the wakefield to produce low relative
energy spread, and completely blow out the electrons left in the bubble after the beam to reach conditions that
preserves emittance. The results have importance for the preparation of AWAKE Run 2~\cite{adli:2016}, and possibly other
applications in the quasi-linear regime.

@@

% 	EA comment: is this part relevant?  Feels out of place.  We repeat this physics below when discussing proton beam SMI.  
%A train of electron drive bunches with a separation $\lambda_{pe}$ and a length $l_{b} \ll \lambda_{pe}$, where the
%plasma wavelength $\lambda_{pe} = 2\pi c/\omega_{pe}$ and the plasma frequency $\omega_{pe} = (n_{0} e^{2} / m_{e}
%\epsilon_{0})^{1/2}$, produces a field $E_{z}$ that increases for each drive bunch~\cite{chen:1985}. A tailing witness
%beam loading the peak accelerating phase of this field will gain energy from the wakefield from all of the drive bunches
%when they are properly phased with respect to the wakefields. Acceleration of an electron witness beam driven by two
%electron drive beams has been demonstrated at Brookhaven National Laboratory~\cite{muggli:2011}.

@@

\begin{figure}[hbt]
    \includegraphics[width=0.99\linewidth,trim={1mm 2mm 1mm 2mm},clip]{figures/figAWAKE}
    \caption{\label{Fig:AWAKER2} A simplified illustration of the experimental setup for AWAKE Run 2. The SPS proton
        beam undergoes self-modulation in the first plasma section. The electron witness beam is injected into the
        accelerating structure, and undergoes acceleration in the second plasma
        section~\cite{berglyd_olsen:2015, adli:2016}}.
\end{figure}

The energy carried by electron drive beams used in previous plasma wakefield experiments have been on the order of
$100\unit{J}$ and the propagation length typically no more than $1\unit{m}$~\cite{blumenfeld:2007, caldwell:2009}.
For high-energy physics application a higher total beam energy is often desired. For instance, the energy of a
high-charge electron beam accelerated to $1\unit{TeV}$ with $1\nexp{10}$ electrons, similar to the beam that could be
produced by the International Linear Collider, is $1.6\unit{kJ}$. Using electron beams as drivers, a large number of
plasma stages is required to reach an energy of a $\unit{kJ}$ for the accelerated beam. However, staging plasma
accelerators without reducing the effective gradient and spoiling the beam quality is challenging
\cite{steinke:2016, lindstrom:2016}.

Proton beams available at CERN carry significantly more energy than available electron beams, $19\unit{kJ}$ for the SPS
beam~\cite{gschwendtner:2016}, allowing for much longer plasma wakefield accelerator stages. The SPS beam is orders of
magnitude longer than the plasma wavelengths needed for such applications, and it does not drive a strong
wake~\cite{gschwendtner:2016}. By letting the proton beam undergo self-modulation before injecting the witness beam into
the accelerating structure, stronger wakefields are excited. The self-modulation is produced by the transverse fields
generated by the beam acting upon itself, causing regions of the beam to rapidly defocus~\cite{kumar:2010}. The
modulation frequency is close to that of the plasma, and produces a train of short proton bunches along the beam axis
with a surrounding halo of defocused particles. This train of bunches resonantly drives wakefields to large amplitudes.

AWAKE at CERN is a proof of concept proton beam driven plasma wakefield accelerator experiment
\cite{awake_collaboration:2014}, currently in its first phase of operation. The experiment uses a $400\unit{GeV}$ proton
beam delivered by the SPS as its driver, and a single $10\unit{m}$ plasma stage with a nominal plasma density of
$7\nexp{14}\unit{cm}^{-3}$~\cite{gschwendtner:2016}. This plasma density corresponds to $\lambda_{pe} = 1.26\unit{mm}$
-- matched to the transverse size of the SPS proton beam such that $k_{pe}\sigma_{x,y,pb} = 1$, where
$k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave number, $\lambda_{pe} = 2\pi c/\omega_{pe}$, and
$\omega_{pe} = \left(n_0e^2/m_e\epsilon_0\right)^{1/2}$ is the plasma electron angular frequency.

The aim of the first phase of the experiment is to demonstrate self-modulation of the proton beam. The aim of the second
phase, in 2018, is to sample the wakefield with a long electron beam $\simeq\lambda_{pe}$. The study presented here has
relevance for Run 2~\cite{adli:2016}, starting in 2021 after the LHC long shutdown 2, and aims to demonstrate
acceleration of a short electron beam, $\ll\lambda_{pe}$, to high energy and with a minimal increase in emittance and
absolute energy spread.

The plans for AWAKE Run 2 proposes to use two plasma sections, illustrated in Fig.~\ref{Fig:AWAKER2}. The first
section of about $4\unit{m}$ is the self-modulation stage where the proton beam undergoes self-modulation without the
electron beam present. The electron witness beam is then injected into the modulated proton beam before section two
where it undergoes acceleration. The self-modulated proton beam does not produce a fully non-linear wakefield, and
therefore not all plasma electrons are evacuated from the plasma bubble. The result is that the focusing force does not
increase linearly with radius and the accelerated beam emittance is not preserved.

@@

The focus in this study is on the beam loading of the wakefields driven by the proton beam. Studies of self-modulated
proton beams show that the proton beam evolves as it propagates through a uniform plasma~\cite{lotov:2011}, but that
small variations in the plasma profile the modulation, and thus the wakefields, may be stabilised over long
distances~\cite{lotov:2011, lotov:2015, caldwell:2011}. To study the witness beam evolution in a stable wake,
independent of the dynamics of the self-modulation, we use a single, non-evolving proton bunch as driver. The proton
beam parameters are chosen so that key features of the wake -- the plasma electron density in the wake and the
longitudinal electric field -- are the same as in the wake of a self-modulated proton beam with AWAKE baseline
parameters~\cite{gschwendtner:2016}. Both the proton beam and the witness beam have Gaussian longitudinal charge
distribution and bi-Gaussian transverse charge distributions.

We have previously studied the beam loading in a proton beam wake using the full particle-in-cell (PIC) code
OSIRIS~\cite{fonseca:2002} with 2D cylindrical-symmetric simulations. The studies~\cite{berglyd_olsen:2015,
berglyd_olsen:2016} primarily looked at beam loading, energy gain and energy spread, as well as different approaches to
creating a stable drive beam structure based on previous self-modulation studies. In order to study the witness beam
emittance evolution we use the recently released open-source version of QuickPIC~\cite{huang:2006, an:2013}. QuickPIC is
a fully relativistic 3D quasi-static PIC code. It does not suffer from the numerical Cherenkov effect that full PIC
codes do~\cite{godfrey:1974,lehe:2013}, making it a well suited tool to study emittance preservation. All simulation
results in this paper were obtained using QuickPIC open-source~\cite{quickpic:web}.

@@

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/plasmaDenTWake}
    \caption{\label{Fig:PlasmaDenTWake} QuickPIC simulations of the initial time step for the single proton drive beam
        and witness beam setup. Plasma electron density is shown in grey with the drive beam (blue) and the witness beam
        (red) superimposed. The line plot indicates the transverse wakefield gradient $\textrm{d}W_{x}/\textrm{d}x$
        where $W_{x} = E_{x} - v_{b} B_{y}$, evaluated along the beam axis. Beams move to the left.}
\end{figure}

@@

The modulation process used in AWAKE does not reach the fully non-linear regime and thus does not produce a bubble void
of plasma electrons. When the SPS beam, containing $3\nexp{11}$ protons~\cite{gschwendtner:2016}, enters the second
plasma section (Fig.~\ref{Fig:AWAKER2}), the peak electric field is expected to be about $500\funit{MV}{m}$. The
plasma electron density is only depleted to around $65\%$ of nominal plasma density at the point where we inject the
electron beam~\cite{awake_collaboration:2016}. The plasma electron density depletion and the peak field is replicated
using a single bunch with $1.46\nexp{10}$ protons ($2.34\unit{nC}$), a length $\sigma_{z} = 40\unit{\mu m}$
($7\unit{kA}$), and a transverse size $\sigma_{x,y,pb} = 200\unit{\mu m}$. The peak density at $0.83\cdot n_{0}$ results
in a quasi-linear wake. To avoid transverse evolution of the proton driver, emulating the stable propagation of the
self-modulated proton beam~\cite{lotov:2011, lotov:2015, caldwell:2011}, we freeze the transverse evolution of the
equivalent proton bunch by increasing the particles mass by six orders of magnitude.

@@

In order to prevent large amplitude oscillations of the witness beam particles, which may cause additional energy spread
as well as emittance growth, we consider a witness beam matched to the plasma density. The matched beam transverse
size~\cite{esarey:1996} is:
\begin{equation}
    \sigma_{x,y,eb}=\left(\frac{2c^{2}\epsilon_{\mathrm{N}}^{2}m_{e}\varepsilon_{0}}{n_{pe}e^{2}\gamma}\right)^{1/4}
    \label{EQ:Matched}
\end{equation}

@@

Beam loading of a short witness beam is sensitive to its position relative to the electric field~\cite{tzoufras:2009}
as well as to its de-phasing with respect to the wakefields at low energy. To eliminate de-phasing of the witness beam,
the initial beam energy is set such that $\gamma_{eb} = \gamma_{pb} = 426.3$, giving an energy of $217\unit{MeV}$. A
lower initial energy is likely to be sufficient for AWAKE Run 2 injection.
% EA: we have studied this to some extent, but unfortunately we do not have a good refence for the last statement,
%     no details in \cite{berglyd_olsen:2015}.

Equation~(\ref{EQ:Matched}) yields a transverse size $\sigma_{x,y,eb}$ of $5.25\unit{\mu m}$, which is narrow compared
to the drive beam $\sigma_{x,y,pb} = 200\unit{\mu m}$. The bunch length was set to $\sigma_{z} = 60\unit{\mu m}$, based
on earlier beam loading studies~\cite{berglyd_olsen:2016}. The charge is adjusted to $100\unit{pC}$ for optimal beam
loading, as discussed in the next section. We refer to the defined drive beam and witness beam parameter set as the
\emph{base case}. Figure~\ref{Fig:PlasmaDenTWake} shows the two beams from a QuickPIC simulation of the initial time
step, for the base case parameters; the proton beam in blue, the trailing electron beam in red, and the plasma electron
density in grey.

@@

The relatively small size of the witness beam puts constraints on the transverse grid cell size and number in the
simulations. We need a small size to resolve the narrow electron beam, and a large number of grid cells to resolve the
much wider proton beam and its wakefields. We use a transverse grid cell size of $1.17\unit{\mu m}$, and of
$2.34\unit{\mu m}$ for the longitudinal grid cells for the simulations presented in section~\ref{S:BL}. The witness beam
was simulated with $16.8\nexp{6}$ non-weighted particles, the drive beam with $2.1\nexp{6}$ non-weighted particles, and
the plasma electrons with $1024 \times 1024$ weighted particles per transverse slice.

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamLoading}
    \caption{\label{Fig:BeamLoading} \textbf{Top plot:} Unloaded longitudinal electric field with no witness beam
        (dashed blue line) and loaded field (whole blue line) along the beam axis. The beam density along the axis for
        both beams are shown in red.
        \textbf{Bottom plot:} Plasma densities along the beam axis for a drive beam with no witness beam (dashed green
        line), witness beam with no drive beam (dash-dotted green line), and both beams present (whole green line).
        $\xi = z - tc$ is the position in the simulation box, moving towards the left. The plots show the initial time
        step.}
\end{figure}

@@

\begin{figure*}[hbt]
    \includegraphics[width=0.495\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamEmittance}
    \includegraphics[width=0.495\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamEmittanceOffset}
    \caption{\label{Fig:BeamEmitt} Beam density in blue along the beam axis for an on-axis beam with respect to the
        drive beam axis (left), and an offset beam (right) with an offset of one $\sigma_{x,eb} = 5.24\unit{\mu m}$ in
        the x-plane -- at four different positions $z$ in the plasma stage. The red lines show a moving window
        calculation of transverse normalised emittance. The moving window calculation uses longitudinal slices of
        $l = 4\cdot\Delta\xi = 9.38\unit{\mu m}$ with a step of $\Delta\xi$. Only slices with more than $100$ macro
        particles have been included. The plasma density profile is included in green, and scaled up by a factor of
        $100$ to be visible. These simulations were run with an LHC energy drive beam of $7\unit{TeV}$}
\end{figure*}

@@

Figure~\ref{Fig:BeamLoading} shows the results of QuickPIC simulations of the initial time step for the base case
parameters. The $E_{z}$-field generated by the proton drive beam is seen as the blue line, shown with and without the
electron beam present. With a proton beam density $n_{pb} \simeq n_{0}$, the wakefields are in the quasi-linear
regime~\cite{rosenzweig:2010}. The dashed green line in the lower part of Fig.~\ref{Fig:BeamLoading} shows that the
on-axis plasma density has a depletion to $67\%$, close to what we see in full scale reference simulations for AWAKE
Run 2~\cite{awake_collaboration:2016}.

@@

The witness beam generates its own wakefield which loads the $E_{z}$-field generated by the drive beam. With an ideally
shaped electron beam charge profile it is possible to optimally load the field in such a way that the accelerating field
is constant along the beam~\cite{katsouleas:1987, tzoufras:2009}. Gaussian beams, as assumed in these studies, cannot
completely flatten the electric field in the tails of the charge distribution, and our base case beam therefore has a
tail in energy both at the front and the back of the beam, as illustrated in Fig.~\ref{Fig:BeamPS}. The bulk of the
beam, however, sees a relatively flat field.
% EA: I rewrote this to the sentence after the comments
%we can acceleratoreHowever, with such a narrow beam when matched beam to the plasma density (equation
%\ref{EQ:MatchedB}), we are limited on how much total charge we can accelerate without reaching charge densities that
%will overload the field.
%  EA: shall we say the below?  The sentence asks for more detailed studies of the dependence of the loading of the transverse fields.
%The amounts of charge required to load the field depends both on the witness bunch length as well as the transverse size.

The initial electron beam density is $n_{eb} \approx 35\cdot n_{0}$. This means that the witness beam's own wakefield
is in the fully non-linear regime, where the space charge force is sufficient to blow out all plasma electrons,
resulting in the formation of an ion column (see Fig.~\ref{Fig:BeamLoading}, bottom). This ion column, as is well
known~\cite{rosenzweig:1991}, provides a linear focusing force on the part the electron beam within the column, and
therefore prevents emittance growth for this part of the beam. This effect is shown for our base case in
Fig.~\ref{Fig:PlasmaDenTWake}. The focusing field has a gradient of $20\unit{kT/m}$ near the beam axis, corresponding to
the matched field gradient.

@@

Figure~\ref{Fig:BeamEmitt} shows the slice emittance along the beam for the base case, sampled after propagating through
$0$, $4$, $40$ and $100\unit{m}$ of plasma. We define emittance of a slice as preserved if the growth is less than
$5\%$, and $\widetilde{Q}$ as the sum charge of the slices for which the emittance is preserved. Simulation results show
that $\widetilde{Q}/Q = 73\%$ of the electron beam longitudinal slices retain their initial emittance after the
propagation in the plasma. Emittance growth mainly occurs in the first few metres, and no significant emittance growth
is observed after this for propagation lengths up to $100\unit{m}$. The head of the beam does not benefit from the full
ion column focusing, but since the proton beam creates a quasi-linear wake, the emittance of the head of the beam still
stabilises after some time. For the $100\unit{m}$ simulation, the drive beam energy was increased to $7\unit{TeV}$ (LHC
energy) to prevent de-phasing, as de-phasing starts to become a significant effect for the SPS beam of $400\unit{GeV}$
after about $50\unit{m}$.
